\documentclass{article}

\usepackage{macros}

\title{Why Geolog?}
\author{Martin Kleppmann}

\begin{document}
\maketitle

\begin{abstract}
    The ARIA Safeguarded AI programme is developing a system called Geolog.
    Other documents explain the details of its design, but leave much of the motivation behind those design decisions implicit.
    This document offers an opinionated justification for why I believe Geolog is important.
    It incorporates insights from others in the SGAI programme, but it's my personal opinion, not a statement for the programme as a whole.
\end{abstract}

\section{The national security case for formal verification}

AI is rapidly advancing in many areas, but one particular area of concern is that AI agents are now able to independently find and exploit large numbers of high-severity security vulnerabilities in widely-used systems~\citep{firefox-vulnerabilities,anthropic-2026}.
As cybersecurity operations are now considered an indispensible element of military strength and deterrence~\citep{DoD-2023-cyber}, we have to assume that these public reports are only the tip of the iceberg, and that nation states have secret research programmes that are stockpiling security vulnerabilities in software, hardware, and protocols for use in a potential future armed conflict.
Just a few years ago, finding and exploiting vulnerabilities in computer systems was expensive~\citep{slayton-2017-cyber}, but this is no longer true.

Cybersecurity generally favours the attacker, who has to find only one vulnerability to compromise a system, whereas the defender has to find and patch \emph{all} vulnerabilities to make their system secure \citep{ryseff-2017-cyberwarfare}.
This fact, together with AI's rapidly growing capability for finding vulnerabilities, worries me a lot.
The theory of \emph{offense-defense balance} \citep{jervis-1978-balance}, which is widely believed in international relations, says that when offensive technologies advance faster than defensive ones, this has a destabilising effect: a country that believes it has a temporary advantage in some offensive technology may decide to attack an opponent in order to exploit that advantage while it lasts.
What matters is not so much whether this offensive advantage is real, but rather whether a potential aggressor \emph{believes} it is sufficient to warrant starting a war \citep{ryseff-2017-cyberwarfare}.

In a time of increasing geopolitical tensions, with commentators warning that we may already be in the foothills of World War 3 \citep{smith-2026-foothills}, there is an urgent need to stabilise the situation as much as possible to reduce the risk of a catastrophic escalation.
(Personally I am less worried about the risk of a superhuman AI deciding to eradicate humanity, and more worried about humans deciding to attack other humans with the help of AI-enabled technology.)
One way we can contribute to stabilisation is through advancing defensive technologies and making them widely available, so that a would-be aggressor is more likely to decide that it is not worth launching an attack in the first place.

In the cybersecurity domain, eliminating vulnerabilities is one of the best defensive measures.
Some types of vulnerability can be avoided by using better tools: for example, using a memory-safe language like Rust instead of C/C++.
However, many critical software systems like operating system kernels or JIT compilers for JavaScript/Wasm in web browsers inherently need to perform unsafe operations (interfacing with hardware devices, or generating machine code from untrusted input and executing it), which cannot be made safe simply by porting to Rust.
Rust also provides only memory safety, but cannot rule out vulnerabilities due to logic bugs, such as incorrect access control checks.

A more comprehensive approach is to use formal verification, i.e.\ to prove that a program conforms to some specification under all possible executions.
This specification can include memory safety properties, but also properties about the logical correctness of the program's security features, and any other properties that are considered important.
These proofs are typically written in a specialised programming language and then checked by a tool such as Lean, Rocq, or Isabelle, which have a small, trusted proof-checking kernel, making it very unlikely that any bugs remain in the verified aspects of the program.

Until recently, formal verification was very expensive, requiring a large time investment from people with highly specialised skills to write the proofs even for small systems.
For example, in 2009, the formally verified seL4 microkernel consisted of 8,700 lines of C code, but proving it correct required 20 person-years and 200,000 lines of Isabelle code~-- or 23 lines of proof and half a person-day for every single line of implementation \citep{Klein2009}.

Fortunately, AI is also getting very good at writing formal proofs; a lot of the current focus is on AI agents for writing proofs in Lean \citep{deMoura2026,Aristotle,LogicalIntelligence,Leanstral,OpenGauss,Chen2025,Ren2025}.
This offers the prospect of formally verifying all of the software on which our critical infrastructure depends in the foreseeable future~-- an idea that would have been ludicrous a few years ago.
And since formal proofs are automatically checked by the kernel, hallucinated proofs are rejected without any human involvement, making formal verification an excellent use case for our current powerful but imperfect LLMs \citep{Kleppmann2025}.

The pervasive use of formally verified computer systems would strengthen the defensive side of cybersecurity, and thereby shift the offense-defense balance towards stability, reducing the risk of armed conflict.
It would also reduce the risk of other harms created by security vulnerabilities, such as the WannaCry ransomware attack that seriously affected the NHS in 2017.

\section{The case for going beyond Lean}

To make this concrete: what would it take to prove that, for example, a version of the Linux kernel or the Firefox web browser is free from security vulnerabilities?
Both of these systems consist of tens of millions of lines of code, making them several orders of magnitude larger than any formal verification project that has been attempted before.
A proof of such scale cannot be accomplished by a single agent~-- it will definitely require many agents working in parallel on different aspects of the proof, for example verifying different modules in the code base, or trying different proof strategies on the same module in parallel to see which one works.

Coordinating between all those agents will require an infrastructure to keep track of which agent is working on what, which theorems have already been proved, which proof strategies have been attempted (including any dead ends, so that they don't get tried again), and so on.
The Categorical Core of SGAI is that infrastructure, and Geolog is the language and data model of the Categorical Core.
Think of Geolog as a database that is optimised for storing large programs and AI-generated proofs about those programs.

In fact, Geolog is a general-purpose database, and not specific to software verification.
Building an automated researcher, as AI firms are now focussing on \citep{Heaven2026}, will require the same kind of infrastructure.
However, I think that examining a concrete use case like ``how could we make the Linux kernel provably secure?'' helps bring clear justifications to design decisions that would otherwise remain abstract and vague.
It also gives us a potential long-term vision to rally around, even if we are still years away from actually being able to verify a system as large as Linux.

AI agents writing formal proofs work best by rapidly iterating on small proof steps, and getting feedback from the proof checker after each step whether it is valid or not~-- much like the interaction that many proof assistants offer to human users, but with the AI providing the `intuition' and guidance on which proof step to try next.
At present, this can be done with Lean by using the Lean Language Server to expose an interactive LSP interface to the prover, and allowing an AI agent to call into this language server via MCP \citep{lean-lsp-mcp}.

That raises the question: why not simply use Lean, given that so much investment in AI-generated proofs is already focussing on Lean?
A core hypothesis of the SGAI program is that Lean will not scale to the size of proofs we are contemplating, because it is fundamentally designed for human-written proofs, and its support for AI-generated proofs is bolted on retroactively.
By starting from a clean slate, we can design Geolog from the ground up to take advantage of the strengths of AI agents.
While this is more work up front, we believe that it will unlock the ability to tackle more ambitious proof projects than will be possible in Lean.

Some of the key differences between Lean and Geolog are:

\begin{itemize}
    \item Lean does not have any mechanism that allows multiple agents to collaborate on a proof, or to backtrack in case they hit a dead end in the proof search.
        The best option is to put the Lean source files in a Git repository and let the agents make Git commits and inspect the Git history in the same way as a human would.
        But concurrent work on the same module would quickly lead to merge conflicts that are messy to resolve at the level of the textual source, and the proof structure has no way of referencing the Git commit history since Lean and Git are two separate systems.
        In contrast, Geolog is a database with built-in version control facilities, including principled ways of merging concurrent work from different agents.
        We are planning to allow the data in the database to reference its own version history, making it possible to record structured information about the meaning of the version history within the system itself.
        For example, the version history could encode the step-by-step evolution of a complex proof, and the database can record which proof rules, axioms, and lemmas were invoked at each step of the proof.

    \item The current integration between Lean and AI agents via MCP and LSP is baroque and slow, since the language server was designed for interaction with humans, not AIs.
        The faster we can make the feedback loop between AI and proof checker, the faster a correct proof can be found.
        With Geolog, we can bring the proof checker as close as possible to the AI model that is driving it: not just on the same machine, but potentially even running on the same GPU as the LLM, to enable extremely fast round-trip times.

    \item Lean uses text files as its representation, i.e.\ a linear sequence of symbols, because this is convenient for humans to enter on a keyboard and convenient to display on a two-dimensional screen for humans to read.
        But text is not amenable for reasoning: in order to prove something about a program, it first has to be parsed into a richer structure such as an abstract syntax tree (AST) or an intermediate representation (IR), and then manipulated at that level.
        AIs are not limited to keyboard input and they can work with multidimensional data better than humans can, and so Geolog skips the textual representation,\footnote{But LLMs are \emph{language models}, I hear you cry, and language is inherently textual!
        Thus, does it really make sense to drop the textual representation?

        Yes, in order for the LLM to interact with Geolog, it will probably output tool call requests as a stream of tokens, and the response from Geolog will go back into the LLM as tokens.
        However, following the principle of small proof steps, these requests and responses will be quite small.
        Rather than feeding an entire module of a program into the LLM as source code text turned into a token stream, it will be much more efficient for Geolog to store the code in a structured form, and to allow the LLM to interactively query it to learn exactly those things that it needs to know about the code.
        The textual representation is then only an incidental detail of the LLM/Geolog interface, and the structured representation stored by Geolog is the primary one.} and allows the AI to interface directly with the AST or IR representation of programs and proofs.
\end{itemize}

Those differences motivate why the SGAI programme is starting afresh with Geolog, and incorporating lessons learnt from Lean and other proof assistants, rather than using the Lean system as it stands today.

\section{The case for a new database}

In addition to being a proof assistant, Geolog is also a new kind of database crossed with a version control system~-- one that is optimised for storing a structured representation of programs and proofs about those programs.
That raises the next question: building a new database is hard (folklore in the databases community says that it takes 10 years to build a new database), so can't we just use an existing one and save ourselves a lot of effort?

There is some related work: for example, Joern CPG \citep{joern-cpg} and Fraunhofer AISEC CPG \citep{fraunhofer-cpg} define ways of mapping code structure onto a graph database such as Neo4j, and Galois MATE CPG \citep{galois-cpg} stores the IR of a program in PostgreSQL.
These tools are already widely used for finding security vulnerabilities through static program analysis.
However, they are not designed to support proofs about programs, and they assume a static program rather than one that is being manipulated by agents.

In the context of AI agents collaborating on proofs, a server-centric database like Neo4j or PostgreSQL is not suitable, for the following reasons:

\begin{itemize}
    \item In order to make the feedback loop for trying proof steps as fast as possible, we cannot have a network round-trip on every access to the database.
        It is essential that each agent can read and write its own local copy of the database, and replicate any writes asynchronously to other agents' replicas without blocking the AI's progress.
        In other words, we need a local-first approach \citep{kleppmann-2019-localfirst}.

    \item In order to allow multiple agents to make progress in parallel without interfering with each other, we may want to let each work for a while on a private \emph{branch} (in the version control sense), and decide later which (if any) branches to merge.
        In case one branch of a proof gets stuck, we want an agent to be able to go back to an older state and try branching off some previous version with a different proof strategy.
        These operations are supported by version control systems, but not by most databases.
\end{itemize}

A version control system like Git is closer to what we need in terms of supporting local reads/writes, branching, and reverting to old versions.
But the downside of most version control systems is that they operate only on text (you can commit binary files to a repository, but the VCS doesn't provide any help with diffing or merging those files).
As discussed previously, we need to operate on a structured data model with well-defined semantics for merging updates from different agents, which Git or similar systems do not provide.
In fact, Git's merge semantics are poorly understood and surprising in various ways \citep{glodny-2025}.

The closest existing system to what we need is Automerge, an embedded database for local-first software with support for Git-like version control that uses CRDTs \citep{shapiro-2025-crdt} to merge branches with well-defined semantics.
However, Automerge's data model is JSON (with a few extensions), which is not well suited for representing programs and proofs.

Fortunately, the Automerge team is part of the SGAI programme.
Our intention is to reuse the infrastructure we have already built for Automerge (physical data layout, versioning mechanisms, compression, replication protocols, access control, sync servers and clients, user interface tooling, \dots), and replace only the JSON data model layer with a new data model that is designed for programs and proofs.
That way, we are starting the 10-year process of building a new database with a head start of several years, while at the same time having the freedom to build a system that is targeted towards the specific needs of the SGAI programme.

\section{The case for a new data model}

But what data model should Geolog use?
The aforementioned CPG projects demonstrate that it is possible to map code ASTs and IR to property graphs and the relational model.
These general-purpose data models have the advantage that they are implemented by existing databases.
But since no existing database meets the needs of SGAI anyway, we also have the opportunity to design a new data model that is better suited to the task.

The relational data model, despite being over half a century old \citep{codd-1970}, has proved extremely versatile and long-lasting.
Almost any data structure, including structured representations of programs and proofs, can be represented in a relational way.
However, the implementation of the relational model in SQL databases has limitations.
In particular, SQL offers only very few types of constraints that an application can declare on a database: there are foreign key constraints, constraints on the values within a single row (check constraints), uniqueness constraints{\dots} and that's it.
There is no way of specifying that the data must have the shape of, say, a linear linked list, or a tree, or a DAG.

In a server-centric database that is not a big problem, because consistency of the database can rely on serializable transactions: if the database starts off in a consistent state, and every transaction transforms the database from one consistent state to another consistent state, then by induction it will remain consistent, even if the consistency property is not explicitly declared as a constraint.
However, in a local-first model there are no serializable transactions (since that would require coordination between concurrent writes, slowing down parallel agents, and it would also rule out branching and merging workflows).
It could then easily happen that two agents make updates that individually preserve consistency, but that break the linearity of the list or the acyclicity of the graph when merged.
Consequently, if we want branching/local-first and also want to ensure that the data has some well-formed structure, we need a more powerful way of declaring constraints than is offered by SQL.

This is the key thing that Geolog provides.
Geolog is a dependently typed programming language that makes it possible to declare a relational schema with database constraints that are vastly more expressive than what SQL provides.
As a user of Geolog you can write \emph{theories} in this programming language, which are definitions of a database schema (data types that can be stored) and \emph{axioms} (logical statements about the data in the database; the database will reject any writes that would violate an axiom).
An \emph{instance} of a theory is a dataset that conforms to the data types and axioms encoded in the theory.

You can think of a Geolog theory as a way of defining a custom data model for a database: it is easy to define a theory of graphs, and then you have a graph database.
But you can also define a theory for an IR to represent programs, and define constraints that ensure it is well-formed.
For example, you can express that a sequence of instructions must appear in a linear order, and that the inputs to one instruction must be the outputs of instructions that appear earlier in that linear order.
You can also express that every type of instruction must have a certain number of inputs, and check that every instruction has the correct number of inputs based on its type.
These are just examples; Geolog is extremely general and flexible in what you can express.

It is even possible to implement a proof checker, analogous to Lean's kernel, using Geolog.
This way, the database will only allow you to write data that is accompanied by a proof that demonstrates why it is true~-- for example, proving that one version of a program is semantically equivalent to another version by referencing an axiom or proof rule that allows a semantic-preserving rewrite of one version into the other version.
Database constraint checking then becomes equivalent to proof checking, and any theorems stored in the database are guaranteed to be true (provided that the proof checker and its axioms are correct).

Now assume that two agents make concurrent updates to a database that individually satisfy the constraints and proof rules, but which violate a constraint when merged.
In this case the database will detect a merge conflict, and ask the AI agent to resolve it.
Due to the precise nature of how constraints are specified in Geolog, resolving this merge conflict is a very different matter from resolving a textual merge conflict in Git: in Geolog, the conflict specifies exactly which constraint is violated by which rows in which tables.
The best way of resolving the conflict will depend on what the agents were trying to achieve; one option might be to simply discard one of the branches.
In some situations, it might be possible to do a CRDT-style automatic merge that ensures all constraints are satisfied in the merged version.
But in general, resolving the merge conflict may not be deterministic, and require the same kind of AI `creativity' or `intuition' that is needed to write proof steps.

Although the Geolog language is very powerful, it is also deliberately restricted compared to other dependently typed programming languages like Agda or Lean.
It only allows the expression of constraints that can be represented in \emph{geometric logic} (hence the name Geolog), a fragment of first-order logic that does not allow negation and restricts the use of implication and quantifiers.
There are several reasons for this choice of logic; the main one is that it guarantees that constraints can be checked efficiently.
In particular, if some new data is written to the database, the system only has to examine the new data that was written, and it doesn't have to go and re-check all of the constraints for all of the existing data~-- in other words, constraints can be checked \emph{incrementally}.
In a large database with frequent writes, that is essential for getting good performance.

These restrictions on the dependent types that can be used in Geolog do not affect the semantic power of the proofs you can write inside a Geolog instance.
The restrictions are only on the logic that is used in the definition of database constraints, not on the logic of the proofs that are stored within the database.
In fact, Geolog says nothing about what logic is used for the proofs within the database~-- it only provides the database, and the proof rules are defined by the authors of Geolog theories.
But with the appropriate theories, anything that can be proved in Lean can also be proved in Geolog.

Also, the restriction on negation applies only to the language for defining constraints, not to queries in general.
It is possible to run arbitrary queries on a database snapshot (an instance of a theory at some point in time)~-- such queries are allowed to use negation, and they could be written in an existing relational query language such as SQL or Datalog.
However, any nonmonotonic queries would not be incremental: that is, if the database changes, you would have to re-run the query to refresh the results.

I have motivated Geolog in the context of reasoning about traditional programs and their IR, but the language is flexible enough to represent a wide variety of different models of computation and reason about them.
It could represent spreadsheets, which are essentially a graph of cells containing formulas that can reference other cells.
It could represent string diagrams, petri nets, probabilistic programming languages, and all sorts of other models.
It is particularly well-suited to diagrammatic forms of reasoning that are widely used in category theory, in contrast to the reasoning based on textual syntax that dominates existing proof assistants.
Hence the database system based on Geolog is called the Categorical Core.

\bibliographystyle{plainnat}
\bibliography{geolog.bib,why.bib}
\end{document}
